---
title: "Statistical Inference Course Project, Part 1"
author: "Eddo W. Hintoso"
output:
  html_document:
    highlights: pygment
    keep_md: yes
    theme: cerulean
    toc: yes
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r, echo = FALSE, results = 'hide'}
library(knitr)
opts_chunk$set(cache = TRUE)
```

****

### Simulate the data
The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. The mean of exponential distribution is `1 / lambda` and the standard deviation is also also `1 / lambda`. Set `lambda = 0.2` for all of the simulations. In this simulation, the distribution of averages of 40 exponentials will be investigated. Note that a thousand or so simulated averages of 40 exponentials are needed.
```{r}
##  set seed for reproducability
set.seed(13)

##  set lambda to 0.2
lambda <- 0.2

##  40 samples
n <- 40

##  1000 simulations
nsim <- 1000

##  simulate
simulatedExponentials <- replicate(nsim, rexp(n, lambda))

##  calculate mean of exponentials
meansExponentials <- apply(simulatedExponentials, 2, mean)
```

****

### Show the sample mean and compare it to the theoretical mean of the distribution.
```{r}
##  distrribution mean
analyticalMean <- mean(meansExponentials)

##  analytical mean
theoreticalMean <- 1 / lambda

##  display results
c("Analytical Mean" = analyticalMean,
  "Theoretical Mean" = theoreticalMean)

##  visualization
library(ggplot2)
ggplot(data = NULL, aes(x = meansExponentials)) +
    geom_histogram(aes(fill = ..count..), binwidth = max(meansExponentials) / n) +
    geom_vline(x = c(analyticalMean,theoreticalMean), col = c("red", "yellow"), show_guide = TRUE) +
    labs(x = "Mean", y = "Frequency") +
    ggtitle("Exponential Function Simulations")
    
```

The red vertical line represents the analytical mean at **`r analyticalMean`**, while the yellow vertical line represents the theoretical mean at **`r theoreticalMean`**. As one can see, the difference between these two means is very small (**`r abs(theoreticalMean - analyticalMean)`**).

****

### Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
```{r}
##  standard deviation and variance of distribution
sdDist <- sd(meansExponentials)
varDist <- sdDist ** 2

##  theoretical standard deviation and variation
sdTheoretical <- (1 / lambda) / sqrt(n)
varTheoretical <- sdTheoretical ** 2

##  display
c("Theoretical Standard Deviation" = sdTheoretical,
  "Simulated Standard Deviation" = sdDist,
  "Theoretical Variance" = varTheoretical,
  "Simulated Variance" = varDist)
```

Standard Deviation of the distribution is **`r sdDist`** with the theoretical SD calculated as **`r sdTheoretical`**. The theoretical variance is calculated as $(\frac{1}{\lambda \sqrt{n}})^2 =$ **`r varTheoretical`**. The actual variance of the distribution is **`r varDist`**.

****

### Show that the distribution is approximately normal.
```{r}
##  compare empirical distribution with normal distribution
qplot(meansExponentials, geom = 'blank') +
    geom_histogram(aes(fill = ..count.., y = ..density..), binwidth = max(meansExponentials) / n) + 
    scale_fill_gradient("Frequency") + 
    geom_line(aes(y = ..density.., color = 'Empirical'), stat = 'density', size = 0.5) + 
    stat_function(fun = dnorm, args = list(mean = analyticalMean, sd = sdDist), aes(color = "Normal"), size = 1) +
    scale_colour_manual(name = 'Density', values = c('red', 'orange')) + 
    labs(x = "Mean", y = "Frequency") +
    ggtitle("Density of Exponential Means")

##  data for qqplot
q <- qqnorm(meansExponentials, plot = FALSE)

##  compare the distribution of averages of 40 exponentials to a normal distribution
ggplot(data = NULL, aes(sample = meansExponentials)) +
    stat_qq(alpha = 0.3, color = "#006699", shape = "circle") +
    geom_smooth(aes(x = q$x, y = q$y), color = "pink", size = 1, method = lm) + 
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
    ggtitle("Normal Q-Q Plot of Exponential Means")
```

Due to the central limit theorem (CLT), the distribution of averages of 40 exponentials is very close to a normal distribution.